---
title: "q02_report"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is the typical volunteer behavior in the first month? In the first 2 months?

How soon do volunteers attend a shift after orientation? How about their 2nd and 3rd shifts?

```{r}
require(dplyr)
require(ggplot2)


master_file <- "./Data/master.csv"
service_file <- "./Data/service.csv"
orient_file <- "./Data/signupsheet.csv"

dfmaster <- read.csv(master_file)
dfService <- read.csv(service_file)
dfOrient <- read.csv(orient_file)



dfOrient$Orientation.Date <- as.Date(dfOrient$Orientation.Date, format = "%m/%d/%y")
dfService$From.date  <- as.Date(dfService$From.date, format = "%m-%d-%Y")

shifts <- dfService[order(dfService$From.date), c("ID", "From.date")]
shifts <- group_by(shifts, ID)
shifts <- data.frame(mutate(shifts, count=seq(n())))

# FIRST SHIFT----

shifts.1 <- shifts[shifts$count == 1,]
df.1 <- merge(dfOrient, shifts.1, by="ID")

df.1$diff <- as.numeric(df.1$From.date - df.1$Orientation.Date)
df.1$diff[df.1$diff < 0] <- NA

# SECOND SHIFT----

shifts.2 <- shifts[shifts$count == 2,]
df.2 <- merge(dfOrient, shifts.2, by="ID")

df.2$diff <- as.numeric(df.2$From.date - df.2$Orientation.Date)
df.2$diff[df.2$diff < 0] <- NA


# THIRD SHIFT----

shifts.3 <- shifts[shifts$count == 3,]
df.3 <- merge(dfOrient, shifts.3, by="ID")

df.3$diff <- as.numeric(df.3$From.date - df.3$Orientation.Date)
df.3$diff[df.3$diff < 0] <- NA


```



```{r, echo=FALSE}
# PLOT SHIFT # RELATIVE TO ORIENTATION DATE


ggplot() +
  geom_bar(data=df.1, aes(x=diff, fill="red"), alpha=0.5) +
  geom_bar(data=df.2, aes(x=diff, fill="blue"), alpha=0.5) +
  geom_bar(data=df.3, aes(x=diff, fill="green"), alpha=0.5) +
  xlim(c(0,60)) +
  ylab("Number of Volunteers") +
  xlab("Number of days after orientation") +
  scale_fill_identity(name="shift number:",
              guide="legend", labels=c("red"="first", "blue"="second", "green"="third"))
```

A large number of volunteers attend a shift within the first 3 weeks of orientation.


## Bonus: What areas are volunteers drawn from?

```{r}
require(zipcode)

df <- read.csv(master_file)
location <- read.csv(orient_file)
location <- location[, c("ID", "Type")]
levels(location$Type) <- c("GF", "GF", "GF", "GF", "GF", "NE", "NE", "PAC", "PAC")
df <- merge(df, location, by = "ID")


ind <- which(colnames(df)=="City" | colnames(df)=="State")
df <- df[, -ind]

df$Zip <- clean.zipcodes(df$Zip)

data(zipcode)

df <- merge(df, zipcode, by.x="Zip", by.y="zip")

#remove data points outside of trisate area
df <- df[df$state == "DE" | df$state == "PA" | df$state == "NJ",]

states <- map_data("state", region=c("pennsylvania", "new jersey", "delaware"))
counties <- map_data("county", region=c("pennsylvania", "new jersey", "delaware"))

#df$hours.grp <- cut(df$Life.hours, summary(df$Life.hours)[c(1,3,6)], include.lowest=T)
df$hours.grp <- cut(df$Life.hours, c(0,4,max(df$Life.hours, na.rm=T)), include.lowest=T)
df.map <- df[!is.na(df$hours.grp),]


```

Plot #1: 
Where do PAWS volunteers reside and how many volunteers from each area recorded more than 4 hours vs 4 or less?
```{r, echo=FALSE}

ggplot() + 
  geom_polygon(data=states, aes(x=long, y=lat, group=group, fill=region), alpha=1) +
  geom_path(data=counties, aes(x=long, y=lat, group=group), color="red") +
  coord_quickmap(xlim=c(-75.7, -74.9), ylim=c(39.7, 40.4)) + #region
  geom_count(data=df.map, aes(x=longitude, y=latitude, color=hours.grp), 
             alpha=0.8, position="jitter") +
  scale_fill_brewer(palette="Greys", direction=-1) +
  scale_color_brewer(palette="Set1")


```

Zoom in on Philly:

```{r, echo=FALSE}

ggplot() + 
  geom_polygon(data=states, aes(x=long, y=lat, group=group, fill=region), alpha=1) +
  geom_path(data=counties, aes(x=long, y=lat, group=group), color="red") +
  coord_quickmap(xlim=c(-75.4, -74.8), ylim=c(39.8, 40.2)) + #zoom
  geom_count(data=df.map, aes(x=longitude, y=latitude, color=hours.grp), 
             alpha=0.8, position="jitter") +
  scale_fill_brewer(palette="Greys", direction=-1) +
  scale_color_brewer(palette="Set1")

```

Plot #2: Which orientation location did volunteers attend?

```{r, echo=FALSE}

df.map <- df[!is.na(df$Type),]
ggplot() + 
  geom_polygon(data=states, aes(x=long, y=lat, group=group, fill=region), alpha=1) +
  geom_path(data=counties, aes(x=long, y=lat, group=group), color="red") +
  coord_quickmap(xlim=c(-75.7, -74.9), ylim=c(39.7, 40.4)) + #region
  geom_count(data=df.map, aes(x=longitude, y=latitude, color=Type), 
             alpha=0.8, position="jitter") +
  scale_fill_brewer(palette="Greys", direction=-1) +
  scale_color_brewer(palette="Set1")

  
```

Zoom in on Philly:

```{r, echo=FALSE}

ggplot() + 
  geom_polygon(data=states, aes(x=long, y=lat, group=group, fill=region), alpha=1) +
  geom_path(data=counties, aes(x=long, y=lat, group=group), color="red") +
  coord_quickmap(xlim=c(-75.4, -74.8), ylim=c(39.8, 40.2)) + #zoom
  geom_count(data=df.map, aes(x=longitude, y=latitude, color=Type), 
             alpha=0.8, position="jitter") +
  scale_fill_brewer(palette="Greys", direction=-1) +
  scale_color_brewer(palette="Set1")

```