---
title: "PAWS Analysis with R-Ladies - Exploring and Manipulating the Data"
author: "Alice Walsh"
date: "3/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs, warning=FALSE}
# This code works when you are using RStudio project. Paths are relative and stuff
suppressPackageStartupMessages(library(tidyverse))
library(ggalluvial)

```

# Question
What is the typical pattern for volunteer disengagement?

I need to summarise the number of volunteers with various characteristics:  

* did orientation, but no shifts
* did orientation, had only 1 shift
* did orientation, had 2+ shifts

What factors are associated with these three groups? location, orientation type, zip code, days to first shift?

For volunteers that do show up, what happens? Lets plot frequencies over the following factors:  

* orientation type
* orientation location
* did a shift
* did a second shift
* shift types

How do we define a "disengaged" volunteer?  

* My idea: Length of time after last shift is > x months (test different x)

# Procedure  
* We have data from orientation sign up - orientation dates
* We have data from PAWS master volunteer database - date for when they were entered and then started
* We have data from volunteer shifts when people logged in and logged out

1. Load in the data
1. Process data to make it analysis-ready
1. Do some data exploration
1. Make a plot to show volunteer patterns
1. Idea: use ggplot2 extension ggalluvial

## Load Data
```{r data_load}
master_file <- '../../Data/master.csv'
service_file <- '../../Data/service.csv'
signup_file <- '../../Data/signupsheet.csv'

master <- read.csv(master_file, quote="", stringsAsFactors = F)
service <- read.csv(service_file, stringsAsFactors = F)
signup <- read.csv(signup_file, stringsAsFactors = F)
```

The `r master_file` has `r nrow(master)` rows.

The `r service_file` has `r nrow(service)` rows.

The `r signup_file` has `r nrow(signup)` rows.

### Basic info on signup table
Signup table has information of the types of orientations.
```{r basic_signup}
knitr::kable(table(signup$Type), caption = "Type")
knitr::kable(table(signup$NOT.Attending), caption = "Not Attending")
knitr::kable(table(signup$Present), caption = "Present")
knitr::kable(table(signup$orientation), caption = "Primary or Secondary Orientation")
```

OK, so I learned that we need to clean up with "Type" fields and remove one row for cancelling on the orientation.

I also created a orientation location that was separate from orientation type.

```{r}
signup_clean <- signup %>%
  filter(NOT.Attending != "CANCELLED") %>%
  mutate(Type = case_when(
    Type == "NE" | Type == "NE + NE" ~ "NE Basic",
    Type == "GF Basic " ~ "GF Basic",
    Type == "GF Basic  + GF WD" ~ "GF Basic + WD",
    Type == "GF Basic + GF WD" ~ "GF Basic + WD",
    TRUE ~ as.character(Type)
  )) %>%
  separate(Type, into = c("orient_loc", "orient_type"), sep = " ", extra = "merge", remove = FALSE) %>%
  select(-NOT.Attending, -Present) %>% # dont need these
  mutate(Orientation.Date = as.Date(Orientation.Date, format = "%m/%d/%y")) %>%
  arrange(Orientation.Date)


with(signup_clean, table(orient_type, orientation, exclude = NULL) )

with(signup_clean, table(orient_loc, orientation, exclude = NULL) )
```

The earliest orientation date was `r min(signup_clean$Orientation.Date)`


Decided to make separate variables for Primary and Secondary orientation (convert to wide).  There were some cases where there were duplicate Secondary orientations for a single ID. Kept the earliest.

```{r}
signup_clean <- reshape(signup_clean,timevar="orientation",idvar="ID",direction="wide")

```

It is known that there are a handful of cases where Primary occured after Secondary orientation - swap those.

```{r}
signup_clean <- signup_clean %>%
  mutate(sec_prim = Orientation.Date.Secondary - Orientation.Date.Primary) %>%
  mutate(new.Date.Primary = case_when(
    sec_prim < 0 ~ Orientation.Date.Secondary,
    TRUE ~ Orientation.Date.Primary
  ),
  new.Date.Secondary = case_when(
    sec_prim < 0 ~Orientation.Date.Primary,
    TRUE ~ Orientation.Date.Secondary
  )) %>%
  mutate(sec_prim_new = new.Date.Secondary - new.Date.Primary) %>%
  select( -Orientation.Date.Primary, -Orientation.Date.Secondary, -sec_prim ) %>%
  rename(Orientation.Date.Primary= new.Date.Primary, Orientation.Date.Secondary= new.Date.Secondary)

```


### Basic info on master table


Can see that again the text fields like city are not clean.  

I cleaned the City variable and converted all zip code to 5 digit level. The zip codes are still messed up as some are not valid 5-digit codes. They might be missing a leading zero.

```{r clean_city}
master_clean <- master %>%
  mutate(city_clean = case_when(
    grepl("Phil|phil", City) ~ "Philadelphia",
    TRUE ~ "other"
  ),
  Zip = case_when(
    grepl("PA|NJ", Zip) ~ "",
    TRUE ~ as.character(Zip)
  )) %>%
  separate(Zip, into="Zip", sep="-", extra = "drop")

```

```{r basic_master}
knitr::kable(table(master_clean$Status, exclude=NULL), caption = "Status")
knitr::kable(table(master_clean$city_clean, exclude=NULL), caption = "City after cleaning")
knitr::kable(table(master_clean$Zip, exclude=NULL), caption = "Zip Codes after cleaning")
```


Look at the date ranges for the dates
```{r master_dates}
master_clean <- master_clean %>%
  mutate(Date.entered = as.Date(Date.entered, format = "%m-%d-%Y"),
         Start.date = as.Date(Start.date, format = "%m-%d-%Y"),
         Stop.date = as.Date(Stop.date, format = "%m-%d-%Y"),
         Orientation.Date.master = as.Date(Orientation.Date, format = "%m-%d-%Y"))

print("Date.entered")
summary(master_clean$Date.entered)
print("Start.date")
summary(master_clean$Start.date)
print("Orientation.date")
summary(master_clean$Orientation.Date.master)

```


### Basic info on service table
Look at min/max dates
```{r basic_service}
service <- service %>%
  mutate(From.date = as.Date(From.date, format = "%m-%d-%Y"),
         To.date = as.Date(To.date, format = "%m-%d-%Y"))

print("From.date")
summary(service$From.date)
print("To.date")
summary(service$To.date)
```

Look at values for factors:
We need to harmonize the site annotation from this table with that in the orientation table.
There are also a ton of assignments that are special events. Let's create one new category for those and lump them all together. 
```{r}
knitr::kable(table(service$Site, exclude=NULL), caption = "Site")
knitr::kable(table(service$Assignment, exclude=NULL), caption = "Assignment")
knitr::kable(table(service$No.Call.No.Show, exclude=NULL), caption = "No.Call.No.Show")
knitr::kable(table(service$Call.Email.to.miss.shift, exclude=NULL), caption = "Call.Email.to.miss.shift")
knitr::kable(table(service$Absence, exclude=NULL), caption = "Absence")

```

```{r cleanup_assignments}
service_clean <- service %>%
  mutate(assignment_group = case_when(
    Assignment == "Cat Care Volunteer - PAC" ~ "Cat Care Volunteer - PAC",
    Assignment == "Cat Care Volunteer - Grant Ave." ~ "Cat Care Volunteer - Grant Ave.",
    Assignment == "Administrative tasks" ~ "Administrative tasks - ",
    Assignment == "Foster Care Administrative Assistant" ~ "Foster Care Administrative Assistant - ",
    Assignment == "Dog Care Volunteer - GF" ~ "Dog Care Volunteer - GF",
    Assignment == "Dog Walking Volunteer - Grant Ave." ~ "Dog Walking Volunteer - Grant Ave.",
    Assignment == "Dog Walking Volunteer - PAC" ~ "Dog Walking Volunteer - PAC",
    Assignment == "Cat Care Volunteer - GF" ~ "Cat Care Volunteer - GF",
    Assignment == "Cat Socialization Team" ~ "Cat Socialization Team - ",
    Assignment == "PAWS South Philly PetSmart" ~ " - PAWS South Philly PetSmart",
    Assignment == "PAC Adoptions Assistant" ~ "Adoptions Assistant - PAC",
    Assignment == "Cat Pawsitive Program Volunteer - Grant Ave. Location" ~ "Cat Pawsitive Program Volunteer - Grant Ave.",
    Assignment == "PAC Dog Mentee Shift (for shadowing only)" ~ "Dog Mentee Shift - PAC",
    Assignment == "PAC Cat Mentee Shift (for shadowing only)" ~ "Cat Mentee Shift - PAC",
    Assignment == "Dog Rescue Volunteer - GF" ~ "Dog Rescue Volunteer - GF",
    Assignment == "Veterinary Tech Volunteer shift - Grant Ave" ~ "Veterinary Tech Volunteer - Grant Ave.",
    Assignment == "PAC Veterinary Tech Volunteer" ~ "Veterinary Tech Volunteer - PAC",
    Assignment == "PAC Orientations" ~ "Orientations - PAC",
    Assignment == "PAC Transport Volunteer" ~ "Transport Volunteer - PAC",
    Assignment == "Grant Mentee Shift (for shadowing only)" ~ "Mentee Shift - Grant Ave.",
    Assignment == "Cat Pawsitive Program Volunteers - OLD CITY" ~ "Cat Pawsitive Program Volunteer - PAC",
    Assignment == "Voicemail Volunteers" ~ "Voicemail Volunteers - ",
    Assignment == "Gray's Ferry Mentee Shift for dog care (for shadowing only)" ~ "Dog Mentee Shift - GF",
    grepl("Mutt Strut", Assignment) ~ "Mutt Strut",
    TRUE ~ "other_event"
  ))


# Should also sep location and task here
service_clean <- service_clean %>%
  separate(assignment_group, into = c("assignment_type", "location"), sep = " - ")

```

```{r}
service_clean %>%
  group_by(assignment_type) %>%
  summarise(Freq = n()) %>%
  arrange(desc(Freq)) %>%
  knitr::kable( caption = "New Assignment Variable")
# knitr::kable(table(service_clean$assignment_type, exclude=NULL), caption = "New Assignment Variable")
```

## Data manipulation



### Filter to after January 1, 2017
How many unique volunteer IDs does this leave in each table?
```{r filter_dates}
service_clean <- service_clean %>%
  filter(From.date >= "2017-01-01")

master_clean <- master_clean %>%
  filter(Date.entered >= "2017-01-01")

length(unique(service_clean$ID))
length(unique(master_clean$ID))
length(unique(signup_clean$ID))
# In all three
length(intersect(signup_clean$ID, intersect(service_clean$ID, master_clean$ID)))
```

98% of volunteers in the signup data are also in the master data. However, only ~55% are in the service data.
Is this because of the way the service data was downloaded?

### Derive service flags
From the service table, derive variables for   

* did 1+ shift?
* did only 1 shift?
* did 2+ shifts?
* did 10+ shifts?

Should we filter out special events?

```{r service_flags, warning=FALSE}

service_summary <- service_clean %>%
  mutate(no_show = Absence > 0 ) %>%
  group_by(ID) %>%
  summarise(total_shifts = sum(no_show==FALSE),
            missed_shifts = sum(no_show==TRUE),
            first_shift_date =  min(From.date[no_show==FALSE], na.rm=T),
            last_recorded_date = max(From.date[no_show==FALSE], na.rm=T),
            time_bw_shifts = as.numeric((max(From.date[no_show==FALSE]) - min(From.date[no_show==FALSE]) )/n()),
            primary_type = names(sort(table(assignment_type), decreasing = T)[1]),
            primary_type_shifts = sort(table(assignment_type), decreasing = T)[1],
            primary_site = names(sort(table(Site), decreasing = T)[1])) 
service_summary$time_bw_shifts[service_summary$total_shifts<=1] <- NA
# Note that this throws a ton of warnings

# Merge with Signup
signup_service <- merge(signup_clean, service_summary, by = "ID", all.x = TRUE) %>%
  mutate(total_shifts = case_when(
    is.na(total_shifts) ~ 0,
    TRUE ~ as.numeric(total_shifts)),
    missed_shifts = case_when(
    is.na(missed_shifts) ~ 0,
    TRUE ~ as.numeric(missed_shifts))) %>%
  mutate(any_shifts = total_shifts != 0,
         only_one_shift = total_shifts == 1,
         twoplus_shifts = total_shifts > 1, 
         tenplus_shifts = total_shifts >= 10)
  
# Add in demographics from master
final_data <- merge(signup_service, select(master_clean, ID, Status, Date.entered, Start.date, city_clean, Zip), by = "ID")
# have some IDs not in master_clean that are in signup_service - these have early date entered

```

## Visualizations
```{r plot}
signup_service_plot <- final_data %>%
  filter(Orientation.Date.Primary <= "2017-08-01") %>%
  group_by(orient_loc.Primary, orient_type.Primary, primary_type, any_shifts) %>%
  summarise(Freq = n()) %>%
  ungroup()


ggplot(data = signup_service_plot,
       aes(axis1 = orient_loc.Primary, axis2 = orient_type.Primary, axis3 = primary_type,
           weight = Freq)) +
  scale_x_discrete(limits = c("Primary Orientation Location", "Type", "primary_type"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = any_shifts)) +
  geom_stratum() + geom_text(stat = "stratum", label.strata = TRUE) +
  theme_minimal() +
  ggtitle("PAWS Volunteer patterns",
          "stratified by whether they did any shifts")
```

OK - This is a start. It needs to be clarified a little.

How long since the last shift? This can help us decide a volunteer disengaged cutoff
```{r calc_diffs}
final_data$total_time <- as.numeric(final_data$last_recorded_date - final_data$Orientation.Date.Primary)
final_data$time_to_first <- as.numeric(final_data$first_shift_date - final_data$Orientation.Date.Primary)
final_data$time_last_to_first <- as.numeric(final_data$last_recorded_date - final_data$first_shift_date)
final_data$days_since_last <- as.numeric(as.Date("2018-03-02") - final_data$last_recorded_date)

```

```{r hist_last}

model_data <- filter(final_data, total_shifts > 0)
hist(as.numeric(model_data$days_since_last), 100,
     main = "Days since last shift for those who did any shifts",
     xlab = "Days")


```

```{r}
median(filter(final_data, total_shifts > 0)$days_since_last)

median(filter(final_data, total_shifts > 0)$time_bw_shifts, na.rm = T)

```

So looking at all the data including recent orientations (where data is censored), the median days since last shift is around a month, and the median time between shifts is about a week.

```{r}
ggplot(filter(final_data, total_shifts > 0), aes(x = Orientation.Date.Primary, y = days_since_last)) +
  geom_point() +
  geom_hline(yintercept = 90, color = "red") + 
  theme_bw()

ggplot(filter(final_data, total_shifts > 0), aes(x = as.Date("2018-03-02") -Orientation.Date.Primary, y = days_since_last)) +
  geom_point() +
  geom_hline(yintercept = 90, color = "red") + 
  geom_abline(slope = 1, intercept = -90, color = "blue") + 
  theme_bw()
```

OK. Based on this plot, I propose to use >90 days (red line) as a cutoff for when a volunteer is disengaged. 
I will restrict to looking at data before a certain date (maybe Sept 01 2017) because we don't know if these volunteers will become disengaged.

## Summary stats
For volunteers that did orientation between Jan 01 2017 and Sept 01 2017:

* How many did orientation, but no shifts?
* How many did orientation, had only 1 shift?
* How many did orientation, had 2+ shifts?

```{r summarise}
end_summary <- final_data  %>%
  filter(Orientation.Date.Primary >= "2017-01-01", Orientation.Date.Primary < "2017-09-01") %>%
  summarise(total_volunteers = n(),
            volunteered_ever = paste( round(100* sum(total_shifts>0)/n(),1) ,"%"),
            volunteered_onlyonce = paste( round(100* sum(total_shifts==1)/n(),1),"%"),
            twoplus = paste( round(100* sum(twoplus_shifts==TRUE)/n(), 1), "%"),
            tenplus = paste( round(100* sum(tenplus_shifts==TRUE)/n(), 1), "%"),
            ave_shifts = round( mean(total_shifts), 2),
            ave_missed_shifts = round( mean(missed_shifts), 2))

knitr::kable(t(end_summary), caption="Overall Summary")

```


# Output

I wrote files for downstream use.

* a Rds with merged data to Data/merged.Rds
* a Rds with cleaned service data to Data/service_clean.Rds

```{r save}
saveRDS(final_data, "../../Data/merged.Rds")
saveRDS(service_clean, "../../Data/service_clean.Rds")
```


# Session Information
