---
title: "PAWS Analysis with R-Ladies - Question 3"
author: "Alice Walsh"
date: "3/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs}
# This code works when you are using RStudio project. Paths are relative and stuff
suppressPackageStartupMessages(library(tidyverse))

```

# Question
What is the typical pattern for volunteer disengagement?

# Procedure  
* We have data from orientation sign up - orientation dates
* We have data from PAWS master volunteer database - date for when they were entered and then started
* We have data from volunteer shifts when people logged in and logged out

1. Load in the data
1. 

## Load Data
```{r data_load}
master_file <- '../../Data/master.csv'
service_file <- '../../Data/service.csv'
signup_file <- '../../Data/signupsheet.csv'

master <- read.csv(master_file, quote="")
service <- read.csv(service_file)
signup <- read.csv(signup_file)
```

The `r master_file` has `r nrow(master)` rows.

The `r service_file` has `r nrow(service)` rows.

The `r signup_file` has `r nrow(signup)` rows.

## Data manipulation
```{r data_manip, warning = FALSE}
# Volunteers have unique IDs
# Lets get Date.entered, Start.date from master and FIRST From.date from service, and Orientation.Date from signup
# Not all IDs in master have ID in signup, not all IDs in service have ID in signup

# merge
dates <- merge(select(master, ID, Date.entered, Start.date, Orientation.Date, Status), 
               filter(signup, orientation == "Primary"), by = "ID", all.x = TRUE)

# originally just wanted to keep first shift, but looked at karla version
# and decided to use her code to make wide :)
# first_service <- service %>%
#   mutate(From.date = as.Date(From.date, format = "%m-%d-%Y")) %>%
#   group_by(ID) %>%
#   summarise(first_service = min(From.date))

service$From.date <- as.Date(service$From.date, format = "%m-%d-%Y")
service <- service[order(service$ID, service$From.date),]
service$shift <- sequence(rle(as.character(service$ID))$lengths)
service <- reshape(service,timevar="shift",idvar="ID",direction="wide")

dates <- merge(dates, select(service, ID, From.date.1, From.date.2, From.date.3, From.date.4, From.date.5), by = "ID", all.x = TRUE)

# Convert to dates
dates <- dates %>%
  mutate(Date.entered = as.Date(Date.entered, format = "%m-%d-%Y"),
         Start.date = as.Date(Start.date, format = "%m-%d-%Y"),
         Orientation.Date.master = as.Date(Orientation.Date.x, format = "%m-%d-%Y"), 
         Orientation.Date.signup = as.Date(Orientation.Date.y, format = "%m/%d/%y")
  ) %>%
  select(ID, Date.entered, Start.date, Orientation.Date.master, Orientation.Date.signup, Status, starts_with("From.date"))
# I had no good ideas to merge two orientation date columns...
dates$Orientation.Date <-  as.Date(apply(dates, 1, function(x) min(x[4], x[5], na.rm = T)))
# One Orientation.Date was 1111-11-11 - change to NA
dates$Orientation.Date[dates$Orientation.Date == "1111-11-11"] <- NA
# Remove individual orientation dates
dates <- select(dates, -Orientation.Date.master, -Orientation.Date.signup)

```



```{r calc_differences}
dates$days_start_orient <- dates$Start.date - dates$Orientation.Date
dates$days_service1_orient <- dates$From.date.1 - dates$Orientation.Date

```


I need to summarise the number of volunteers with various characteristics, such as no orientation, took orientation, but never volunteered, etc.

# Output

I wrote the Rds with merged data to Data/merged.Rds for downstream use.
```{r save}
saveRDS(dates, "../../Data/merged_dates.Rds")
```


# Session Information
