---
title: "PAWS Analysis with R-Ladies - Question 3"
author: "Alice Walsh"
date: "3/16/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs}
# This code works when you are using RStudio project. Paths are relative and stuff
suppressPackageStartupMessages(library(tidyverse))
library(ggalluvial)

```

# Question
What is the typical pattern for volunteer disengagement?

I need to summarise the number of volunteers with various characteristics:  
* did orientation, but no shifts
* did orientation, had only 1 shift
* did orientation, had 2+ shifts
What factors are associated with these three groups? location, orientation type, zip code, days to first shift?

For volunteers that do show up, what happens? Lets plot frequencies over the following factors:  
* orientation type
* orientation location
* did a shift
* did a second shift
* shift types

# Procedure  
* We have data from orientation sign up - orientation dates
* We have data from PAWS master volunteer database - date for when they were entered and then started
* We have data from volunteer shifts when people logged in and logged out

1. Load in the data
1. Process data to make it analysis-ready
1. Do some data exploration
1. Make a plot to show volunteer patterns
1. Idea: use ggplot2 extension ggalluvial

## Load Data
```{r data_load}
master_file <- '../../Data/master.csv'
service_file <- '../../Data/service.csv'
signup_file <- '../../Data/signupsheet.csv'

master <- read.csv(master_file, quote="", stringsAsFactors = F)
service <- read.csv(service_file, stringsAsFactors = F)
signup <- read.csv(signup_file, stringsAsFactors = F)
```

The `r master_file` has `r nrow(master)` rows.

The `r service_file` has `r nrow(service)` rows.

The `r signup_file` has `r nrow(signup)` rows.

### Basic info on signup table
Signup table has information of the types of orientations.
```{r basic_signup}
knitr::kable(table(signup$Type), caption = "Type")
knitr::kable(table(signup$NOT.Attending), caption = "Not Attending")
knitr::kable(table(signup$Present), caption = "Present")
knitr::kable(table(signup$orientation), caption = "Primary or Secondary Orientation")
```

OK, so I learned that we need to clean up with "Type" fields and remove one row for cancelling on the orientation.

I also created a orientation location that was separate from orientation type.

```{r}
signup_clean <- signup %>%
  filter(NOT.Attending != "CANCELLED") %>%
  mutate(Type = case_when(
    Type == "NE" | Type == "NE + NE" ~ "NE Basic",
    Type == "GF Basic " ~ "GF Basic",
    Type == "GF Basic  + GF WD" ~ "GF Basic + WD",
    Type == "GF Basic + GF WD" ~ "GF Basic + WD",
    TRUE ~ as.character(Type)
  )) %>%
  separate(Type, into = c("orient_loc", "orient_type"), sep = " ", extra = "merge", remove = FALSE) %>%
  select(-NOT.Attending, -Present) %>% # dont need these
  mutate(Orientation.Date = as.Date(Orientation.Date, format = "%m/%d/%y")) %>%
  arrange(Orientation.Date)


with(signup_clean, table(orient_type, orientation, exclude = NULL) )

with(signup_clean, table(orient_loc, orientation, exclude = NULL) )
```

The earliest orientation date was `r min(signup_clean$Orientation.Date)`


Decided to make separate variables for Primary and Secondary orientation (convert to wide).  There were some cases where there were duplicate Secondary orientations for a single ID. Kept the earliest.

```{r}
signup_clean <- reshape(signup_clean,timevar="orientation",idvar="ID",direction="wide")

```

### Basic info on master table

```{r basic_master}
knitr::kable(table(master$Status, exclude=NULL), caption = "Status")
knitr::kable(table(master$City, exclude=NULL), caption = "City")
knitr::kable(table(master$Zip, exclude=NULL), caption = "Zip Codes")


```

Can see that again the text fields like city are not clean. I did not yeat clean them.

Look at the date ranges for the dates
```{r master_dates}
master <- master %>%
  mutate(Date.entered = as.Date(Date.entered, format = "%m-%d-%Y"),
         Start.date = as.Date(Start.date, format = "%m-%d-%Y"),
         Stop.date = as.Date(Stop.date, format = "%m-%d-%Y"),
         Orientation.Date.master = as.Date(Orientation.Date, format = "%m-%d-%Y"))

print("Date.entered")
summary(master$Date.entered)
print("Start.date")
summary(master$Start.date)
print("Orientation.date")
summary(master$Orientation.Date.master)

```


### Basic info on service table
Look at min/max dates
```{r basic_service}
service <- service %>%
  mutate(From.date = as.Date(From.date, format = "%m-%d-%Y"),
         To.date = as.Date(To.date, format = "%m-%d-%Y"))

print("From.date")
summary(service$From.date)
print("To.date")
summary(service$To.date)
```

Look at values for factors:
We need to harmonize the site annotation from this table with that in the orientation table.
There are also a ton of assignments that are special events. Let's create one new category for those and lump them all together. 
```{r}
knitr::kable(table(service$Site, exclude=NULL), caption = "Site")
knitr::kable(table(service$Assignment, exclude=NULL), caption = "Assignment")
knitr::kable(table(service$No.Call.No.Show, exclude=NULL), caption = "No.Call.No.Show")
knitr::kable(table(service$Call.Email.to.miss.shift, exclude=NULL), caption = "Call.Email.to.miss.shift")
knitr::kable(table(service$Absence, exclude=NULL), caption = "Absence")

```

```{r}
service_clean <- service %>%
  mutate(assignment_group = case_when(
    Assignment == "Cat Care Volunteer - PAC" ~ "Cat Care Volunteer - PAC",
    Assignment == "Cat Care Volunteer - Grant Ave." ~ "Cat Care Volunteer - Grant Ave.",
    Assignment == "Administrative tasks" ~ "Administrative tasks - ",
    Assignment == "Dog Care Volunteer - GF" ~ "Dog Care Volunteer - GF",
    Assignment == "Dog Walking Volunteer - Grant Ave." ~ "Dog Walking Volunteer - Grant Ave.",
    Assignment == "Cat Care Volunteer - GF" ~ "Cat Care Volunteer - GF",
    Assignment == "Cat Socialization Team" ~ "Cat Socialization Team - ",
    Assignment == "PAWS South Philly PetSmart" ~ " - PAWS South Philly PetSmart",
    Assignment == "PAC Adoptions Assistant" ~ "Adoptions Assistant - PAC",
    Assignment == "Cat Pawsitive Program Volunteer - Grant Ave. Location" ~ "Cat Pawsitive Program Volunteer - Grant Ave.",
    grepl("Mutt Strut", Assignment) ~ "Mutt Strut",
    TRUE ~ "other_event"
  ))

# Should also sep location and task here
service_clean <- service_clean %>%
  separate(assignment_group, into = c("assignment_type", "location"), sep = " - ")

```

```{r}
knitr::kable(table(service_clean$assignment_type, exclude=NULL), caption = "New Assignment Variable")
```

## Data manipulation



### Filter to after January 1, 2017
How many unique volunteer IDs does this leave in each table?
```{r filter_dates}
service_clean <- service_clean %>%
  filter(From.date >= "2017-01-01")

master_clean <- master %>%
  filter(Date.entered >= "2017-01-01")

length(unique(service_clean$ID))
length(unique(master_clean$ID))
length(unique(signup_clean$ID))
# In all three
length(intersect(signup_clean$ID, intersect(service_clean$ID, master_clean$ID)))
```

### Derive service flags
From the service table, derive variables for   

* did 1+ shift?
* did only 1 shift?
* did 2+ shifts?
* did 3+ shifts?

Should we filter out special events?

```{r service_flags}
service_summary <- service_clean %>%
  group_by(ID) %>%
  summarise(total_shifts = n())

# Merge with Signup
signup_service <- merge(signup_clean, service_summary, by = "ID", all.x = TRUE) %>%
  mutate(total_shifts = case_when(
    is.na(total_shifts) ~ 0,
    TRUE ~ as.numeric(total_shifts)
  )) %>%
  mutate(any_shifts = total_shifts != 0,
         only_one_shift = total_shifts == 1,
         twoplus_shifts = total_shifts > 1, 
         tenplus_shifts = total_shifts >= 10)
  
```

```{r plot}
signup_service_plot <- signup_service %>%
  group_by(orient_loc.Primary, orient_type.Primary, twoplus_shifts, any_shifts) %>%
  summarise(Freq = n()) %>%
  ungroup()


ggplot(data = signup_service_plot,
       aes(axis1 = orient_loc.Primary, axis2 = orient_type.Primary, axis3 = twoplus_shifts,
           weight = Freq)) +
  scale_x_discrete(limits = c("Primary Orientation Location", "Type", "Did 2+ shifts"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = any_shifts)) +
  geom_stratum() + geom_text(stat = "stratum", label.strata = TRUE) +
  theme_minimal() +
  ggtitle("PAWS Volunteer patterns",
          "stratified by whether they did any shifts")
```

OK - This is a start. It needs to be clarified a little.

# Output

I wrote the Rds with merged data to Data/merged.Rds for downstream use.
```{r save}
# saveRDS(signup_service, "../../Data/merged.Rds")
```


# Session Information
