---
title: "PAWS Analysis with R-Ladies - Question 3 additional analysis"
author: "Alice Walsh"
date: "3/23/2018"
output: html_document
---

```{r, echo=FALSE, include=FALSE}
rm(list = ls()) # can modify this if we are passing parameters into rmarkdown, etc.
set.seed(1234)
options(stringsAsFactors = FALSE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs, warning=FALSE, message=FALSE}
# This code works when you are using RStudio project. Paths are relative and stuff
library(survival)
library(survminer)
library(RColorBrewer)
suppressPackageStartupMessages(library(tidyverse))

```

```{r additional_setup, include=FALSE}
# I like to change the default color palette
scale_color_discrete <- function(...) {scale_color_manual(..., values = brewer.pal(n=9, name = "Set1"))}
scale_fill_discrete <- function(...) {scale_fill_manual(..., values = brewer.pal(n=9, name = "Set1"))}

```

# Question
What is the typical pattern for volunteer disengagement?

I need to summarise the number of volunteers with various characteristics:  

* did orientation, but no shifts
* did orientation, had only 1 shift
* did orientation, had 2+ shifts

What factors are associated with these three groups? location, orientation type, zip code, days to first shift?

For volunteers that do show up, what happens? Lets plot frequencies over the following factors:  

* orientation type
* orientation location
* did a shift
* did a second shift
* shift types

How do we define a "disengaged" volunteer?  

* My idea: Length of time after last shift is > x months (test different x)

# Procedure  

1. Load in the data
1. Analyze

## Load Data
```{r}
final_data <- readRDS("../../Data/merged.Rds")
```

# Analysis
Let's try time-to-event analysis. Here we will look at data for volunteers who did orientation between Jan 1, 2017 and August 1, 2017. For this analysis, only volunteers that did at least one shift are considered and a volunteer is defined as "disengaged" if greater than 90 days passes without volunteering.

```{r calc_diffs}
final_data$total_time <- final_data$last_recorded_date - final_data$Orientation.Date.Primary
final_data$days_since_last <- as.numeric(as.Date("2018-03-02") - final_data$last_recorded_date)
final_data$censor <- 0
final_data$censor[final_data$days_since_last > 90] <- 1 
```

```{r kaplan-meier}
# Show curves for last service - need survival and survminer loaded

model_data <- filter(final_data, any_shifts > 0, Orientation.Date.Primary <= "2017-08-01")
surv.model <- Surv(model_data$total_time, model_data$censor)
surv.fit <- survfit(surv.model ~ only_one_shift, data = model_data)
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = TRUE,
           ylab = "Volunteer Engagement Prob",
           xlab = "Time (days)",
           title = "Compare whether only did one shift",
           legend.title = "", legend.labs = levels(factor(model_data$only_one_shift)))  


model_data <- filter(final_data, any_shifts > 0,
                     Orientation.Date.Primary <= "2017-08-01",
                     primary_type %in% c("Cat Care Volunteer", "Dog Care Volunteer","other_event"))
surv.model <- Surv(model_data$total_time, model_data$censor)
surv.fit <- survfit(surv.model ~ primary_type, data = model_data)
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = FALSE,
           ylab = "Volunteer Engagement Prob", xlab = "Time (days)",
           title = "Compare primary assignments",
           legend.title = "", legend.labs = levels(factor(model_data$primary_type)))  
  

model_data <- filter(final_data, any_shifts > 0, Orientation.Date.Primary <= "2017-08-01")
surv.model <- Surv(model_data$total_time, model_data$censor)
surv.fit <- survfit(surv.model ~ orient_loc.Primary, data = model_data)
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = FALSE,
           ylab = "Volunteer Engagement Prob", xlab = "Time (days)",
           title = "Compare primary orientation location",
           legend.title = "", legend.labs = levels(factor(model_data$orient_loc.Primary))) 
```


