---
title: "PAWS Analysis with R-Ladies - Question 3 additional analysis"
author: "Alice Walsh"
date: "3/23/2018"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs, warning=FALSE, message=FALSE}
# This code works when you are using RStudio project. Paths are relative and stuff
suppressPackageStartupMessages(library(tidyverse))
library(survival)
library(survminer)
```

# Question
What is the typical pattern for volunteer disengagement?

I need to summarise the number of volunteers with various characteristics:  

* did orientation, but no shifts
* did orientation, had only 1 shift
* did orientation, had 2+ shifts

What factors are associated with these three groups? location, orientation type, zip code, days to first shift?

For volunteers that do show up, what happens? Lets plot frequencies over the following factors:  

* orientation type
* orientation location
* did a shift
* did a second shift
* shift types

How do we define a "disengaged" volunteer?  

* My idea: Length of time after last shift is > x months (test different x)

# Procedure  

1. Load in the data
1. Analyze

## Load Data
```{r}
final_data <- readRDS("../../Data/merged.Rds")
```

# Analysis
Let's try time-to-event analysis - like survival curves used in cancer studies.

```{r kaplan-meier}
# Show curves for last service - need survival and survminer loaded
final_data$total_time <- final_data$last_recorded_date - final_data$Orientation.Date.Primary
final_data$days_since_last <- as.numeric(as.Date("2018-03-02") - final_data$last_recorded_date)
final_data$censor <- 0
final_data$censor[final_data$days_since_last > 90] <- 1 
  
  
model_data <- filter(final_data, any_shifts > 0)
surv.model <- Surv(model_data$total_time, model_data$censor)
surv.fit <- survfit(surv.model ~ only_one_shift, data = model_data)
ggsurvplot(surv.fit, ylab = "Volunteer Engagment Prob", xlab = "Time (days)")  


model_data <- filter(final_data, any_shifts > 0 & primary_type %in% c("Cat Care Volunteer", "Dog Care Volunteer","other_event"))
surv.model <- Surv(model_data$total_time, model_data$censor)
surv.fit <- survfit(surv.model ~ primary_type, data = model_data)
ggsurvplot(surv.fit, ylab = "Volunteer Engagment Prob", xlab = "Time (days)")  
  
```


