---
title: "PAWS Analysis with R-Ladies - Question 3 additional analysis"
author: "Alice Walsh (updated by KF)"
date: "8/20/2018"
output:
  pdf_document: default
  html_document: default
---

```{r, echo=FALSE, include=FALSE}
rm(list = ls()) # can modify this if we are passing parameters into rmarkdown, etc.
set.seed(1234)
options(stringsAsFactors = FALSE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs, warning=FALSE, message=FALSE}
# This code works when you are using RStudio project. Paths are relative and stuff
library(survival)
library(survminer)
library(RColorBrewer)
suppressPackageStartupMessages(library(tidyverse))

```

```{r additional_setup, include=FALSE}
# I like to change the default color palette - doesn't apply to ggsurvplot for some reason?
scale_color_discrete <- function(...) {scale_color_manual(..., values = brewer.pal(n=9, name = "Set1"))}
scale_fill_discrete <- function(...) {scale_fill_manual(..., values = brewer.pal(n=9, name = "Set1"))}

```

# Question
What is the typical pattern for volunteer disengagement?

I need to summarise the number of volunteers with various characteristics:  

* did orientation, but no shifts
* did orientation, had only 1 shift
* did orientation, had 2+ shifts

What factors are associated with these three groups? location, orientation type, zip code, days to first shift?

For volunteers that do show up, what happens? Lets plot frequencies over the following factors:  

* orientation type
* orientation location
* did a shift
* did a second shift
* shift types

How do we define a "disengaged" volunteer?  

* My idea: Length of time after last shift is > x months (test different x)

# Procedure  

1. Load in the data
1. Analyze

## Load Data
```{r}
final_data <- readRDS("../../../Data/merged.Rds")
```

# Analysis
Let's try time-to-event analysis. Here we will look at data for volunteers who did orientation between Jan 1, 2018 and May 1, 2018. For this analysis, only volunteers that did at least one shift are considered and a volunteer is defined as "disengaged" if greater than 90 days passes without volunteering.

```{r calc_disengage}
final_data$event <- 0
final_data$event[final_data$days_since_last > 90] <- 1 

# create factor for whether they reported for shift within 14 days
final_data$first_shift_group <- case_when(
  as.numeric(final_data$time_to_first) <= 14 ~ "<= 14 days",
  as.numeric(final_data$time_to_first) <= 60 ~ "14 to 60 days",
  as.numeric(final_data$time_to_first) > 60 ~ "> 60 days"
)
final_data$first_shift_group <- factor(final_data$first_shift_group,
                                       levels = c("<= 14 days", "14 to 60 days", "> 60 days"))

final_data$timebw_shift_group <- case_when(
  as.numeric(final_data$time_bw_median) <= 7 ~ "<= 7 days",
  as.numeric(final_data$time_bw_median) <= 14 ~ "7 to 14 days",
  as.numeric(final_data$time_bw_median) > 14 ~ "> 14 days"
)
final_data$timebw_shift_group <- factor(final_data$timebw_shift_group, 
                                        levels = c("<= 7 days", "7 to 14 days", "> 14 days"))

final_data %>%
  filter(Orientation.Date.Primary <= "2018-05-01", total_shifts > 0) %>%
  group_by(first_shift_group) %>%
  summarise(mean_shifts = mean(total_shifts),
            mean_missed_shifts = mean(missed_shifts),
            freq = n()) %>%
  knitr::kable(caption = "Summary of time orientation to first shift")

final_data %>%
  filter(Orientation.Date.Primary <= "2018-05-01", total_shifts > 0 ) %>%
  group_by(timebw_shift_group) %>%
  summarise(mean_shifts = mean(total_shifts),
            mean_missed_shifts = mean(missed_shifts),
            freq = n()) %>%
  knitr::kable(caption = "Summary of time between shifts")
```


## Plot volunteers time between first shift and disengagement (time_last_to_first)

```{r kaplan-meier1}
model_data <- filter(final_data, total_shifts > 1, Orientation.Date.Primary <= "2018-05-01")
surv.model <- Surv(model_data$time_last_to_first, model_data$event)

surv.fit <- survfit(surv.model ~ first_shift_group, data = model_data)
surv.fit
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = TRUE, cumevents = TRUE,
           ylab = "Volunteer Engagement Prob",
           palette = brewer.pal(n=9, "Set1"),
           xlab = "Time From First to Last Shift (days)",
           title = "Compare how soon after orientation volunteered",
           legend.title = "", legend.labs = levels(factor(model_data$first_shift_group)))  

surv.fit <- survfit(surv.model ~ timebw_shift_group, data = model_data)
surv.fit
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = TRUE, cumevents = TRUE,
           ylab = "Volunteer Engagement Prob",
           palette = brewer.pal(n=9, "Set1"),
           xlab = "Time From First to Last Shift (days)",
           title = "Compare median time between shifts",
           legend.title = "", legend.labs = levels(factor(model_data$timebw_shift_group)))  

```

We don't see a strong trend that volunteers leave sooner if they tend to volunteer more frequently or if they first volunteer soon or later after they complete orientation.

```{r kaplan-meier}
# Could plot either total_time - time from orientation to last shift
# or time_last_to_first - time from first shift to last shift

model_data <- filter(final_data, total_shifts > 0, Orientation.Date.Primary <= "2018-05-01")
surv.model <- Surv(model_data$total_time, model_data$event)
surv.fit <- survfit(surv.model ~ only_one_shift, data = model_data)
surv.fit
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = TRUE, cumevents = TRUE,
           ylab = "Volunteer Engagement Prob",
           palette = brewer.pal(n=9, "Set1"),
           xlab = "Time From Orientation to Last Shift (days)",
           title = "Compare whether only did one shift",
           legend.title = "", legend.labs = levels(factor(model_data$only_one_shift)))  

# top assignments:
top_assign <- names(sort(table(final_data$primary_type),decreasing = T))[1:3]
model_data <- filter(final_data, total_shifts > 1,
                     Orientation.Date.Primary <= "2018-05-01",
                     primary_type %in% top_assign)

surv.model <- Surv(model_data$time_last_to_first, model_data$event)
surv.fit <- survfit(surv.model ~ primary_type, data = model_data)
surv.fit
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = FALSE, cumevents = TRUE,
           palette = brewer.pal(n=9, "Set1"),
           ylab = "Volunteer Engagement Prob", xlab = "Time From First to Last Shift (days)",
           title = "Compare primary assignments",
           legend.title = "", legend.labs = levels(factor(model_data$primary_type)))  
  

model_data <- filter(final_data, total_shifts > 1, Orientation.Date.Primary <= "2018-05-01")
surv.model <- Surv(model_data$time_last_to_first, model_data$event)
surv.fit <- survfit(surv.model ~ orient_loc.Primary, data = model_data)
surv.fit
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = FALSE, cumevents = TRUE,
           palette = brewer.pal(n=9, "Set1"),
           ylab = "Volunteer Engagement Prob", xlab = "Time From First to Last Shift (days)",
           title = "Compare primary orientation location",
           legend.title = "", legend.labs = levels(factor(model_data$orient_loc.Primary))) 


model_data <- filter(final_data, total_shifts > 1, Orientation.Date.Primary <= "2018-05-01")
surv.model <- Surv(model_data$time_last_to_first, model_data$event)
surv.fit <- survfit(surv.model ~ city_clean, data = model_data)
surv.fit
ggsurvplot(surv.fit, data = model_data,
           surv.median.line = "hv", 
           pval = TRUE, conf.int = FALSE, cumevents = TRUE,
           palette = brewer.pal(n=9, "Set1"),
           ylab = "Volunteer Engagement Prob", xlab = "Time From First to Last Shift (days)",
           title = "Compare primary orientation location",
           legend.title = "", legend.labs = levels(factor(model_data$city_clean))) 
```

There is not a clear trend for orientation location or whether someone lives in philadelphia or outside philadelphia.

There is a trend that volunteers that typically volunteer as dog walking volunteers stay engaged longer than cat care of dog care volunteers. Cat care is at all three sites, dog care only at Grays Ferry and dog walking is either PAC or NE.

